---
title: "Homework2"
author: "Zhuodiao Kuang"
date: "`r Sys.Date()`"
output: github_document
---

```{r Load libraries, message = FALSE}
library(tidyverse)
library(dplyr)
library(readxl)
library(knitr)
```

```{r setup, echo = FALSE, cache = FALSE, message = FALSE}
rm(list = ls())

## Global options
opts_chunk$set(echo      = TRUE,
	             cache     = TRUE,
               prompt    = FALSE,
               tidy      = FALSE,
               comment   = NA,
               message   = FALSE,
               warning   = FALSE,
               dpi       = 150,
               fig.width = 4, 
               fig.asp   = 5/7,
               fig.align = "center")

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```






# Problem 1

We clean the 538 `pols` data, which provides information on the number of national politicians who are democratic or republican at any given time. There are some values for which `prez_gop` is `2` -- these are months in which Ford became President following Nixon's resignation. In the new `president` variable created as part of our data cleaning, we code these as `rep`` (same as values when `prez_gop` is `1`). 

prez_gop: indicator of whether the president was republican on the associated date (1 = yes, 0 = no)
gov_gop: the number of republican governors on the associated date
sen_gop: the number of republican senators on the associated date
rep_gop: the number of republican representatives on the associated date
prez_dem: indicator of whether the president was democratic on the associated date (1 = yes, 0 = no)
gov_dem: the number of democratic governors on the associated date
sen_dem: the number of democratic senators on the associated date
rep_dem: the number of democratic representatives on the associated date

# First step

```{r clean_538_pols}
month_df = 
  tibble(
    month_num = 1:12,
    month_abb = month.abb,
    month = month.name
  )

pm_df = 
  read_csv("./data/pols-month.csv") |>
  separate(mon, into = c("year", "month_num", "day"), convert = TRUE) |>
  mutate(
    president = recode(prez_gop, "0" = "dem", "1" = "rep", "2" = "rep")) |>
  left_join(x = _, y = month_df) |> 
  select(year, month, everything(), -day, -starts_with("prez"))  
```


# Second step

```{r load `snp.csv`}
snp_df = 
  read_csv("./data/snp.csv") |>
  separate(date, into = c("month", "day", "year"), convert = TRUE) |>
  arrange(year, month) |>
  mutate(month = month.name[month],year = ifelse(year>=50, 1900+year, 2000+year)) |> 
  select(year, month, close) 

```

# Third step

Finally, we tidy the `unemployment` data so that it can be merged with the `pols` and `snp` datasets.

```{r clean_unemp}
unemployment = 
  read_csv("./data/unemployment.csv") |>
  rename(year = Year) |>
  pivot_longer(
    Jan:Dec, 
    names_to = "month_abb",
    values_to = "unemployment"
  ) |> 
  left_join(x = _, y = month_df) |> 
  select(year, month, unemployment)
```


# Last step


Now we merge the three datasets!

```{r merge}
data= 
  left_join(pm_df, snp_df) |>
  left_join(x = _, y = unemployment)

str(data)

```

So the pols_month dataset contains the information of presidents, governors, senators, and representatives from Republican and Democratic during different periods. 

The snp dataset contains the Standard & Poor’s stock market index during different periods. 

The unemployment dataset contains the unemployment rate in the US during different periods. 

The period of the dataset is from 1947 to 2015. The size of the merged dataset is 822 x 12.


# Problem 2

### Read and clean the Mr. Trash Wheel sheet

```{r}
# load the dataset, change the format of date, recalculate homes_powered
mr_trash_wheel_df = 
  readxl::read_excel("./data/202309TWCD.xlsx", sheet = "Mr. Trash Wheel", range = "A2:M586")|>
  janitor::clean_names()|>
  drop_na(dumpster) |> # omit rows that do not include dumpster-specific data
  mutate(
    date = substr(date, nchar(date) - 1, nchar(date)),
    year = as.character(year), # alter the type of year
    homes_powered = (weight_tons*500/30),
    name = "Mr_trash_wheel")|># add a column to specify
  rename(day = date)|>
  select(dumpster,name,month,year,day,everything())
head(mr_trash_wheel_df)
```

### Use a similar process to import, clean, and organize the data for Professor Trash Wheel and Gwynnda.

```{r}
prof_trash_wheel_df = 
  readxl::read_excel("./data/202309TWCD.xlsx", sheet = "Professor Trash Wheel", range = "A2:M108")|> 
  janitor::clean_names() |> # use reasonable variable names
  drop_na(dumpster) |> # omit rows that do not include dumpster-specific data
  mutate(homes_powered = weight_tons * 500 / 30,
         date = substr(date, nchar(date) - 1, nchar(date)),
         year = as.character(year), # alter the type of year
         name = "Professory")|>
  rename(day = date)|>
  select(dumpster,name,month,year,day,everything())
head(prof_trash_wheel_df)
```


```{r}
gwy_trash_wheel_df = 
  readxl::read_excel("./data/202309TWCD.xlsx", sheet = "Gwynnda Trash Wheel", range = "A2:J157")|> 
  janitor::clean_names() |> # use reasonable variable names
  drop_na(dumpster) |> # omit rows that do not include dumpster-specific data
  mutate(homes_powered = weight_tons * 500 / 30,
         date = substr(date, nchar(date) - 1, nchar(date)),
         year = as.character(year), # alter the type of year
         name = "Gwynnda")|>
  rename(day = date)|>
  select(dumpster,name,month,year,day,everything())
head(gwy_trash_wheel_df)
```


### Then, combine the three datasets.

```{r}
trash_wheel_df = 
  mr_trash_wheel_df |> 
  full_join(prof_trash_wheel_df) |> 
  full_join(gwy_trash_wheel_df) |> 
  relocate(name) # put the name of the trash wheel front
head(trash_wheel_df )
```


The three given data sets are all related to the total weight and volume of the trash, the number of different waste products, the number of the dumpster, and the number of households the trash equates to in terms of electricity on a given date.  

Specifically, the cleaned "Mr. Trash Wheel" data set contains `r nrow(mr_trash_wheel_df)` observations of `r ncol(mr_trash_wheel_df)` variables from `r paste(mr_trash_wheel_df$month[1], min(mr_trash_wheel_df$year), sep = ", ")` to `r paste(mr_trash_wheel_df$month[nrow(mr_trash_wheel_df)], max(mr_trash_wheel_df$year), sep = ", ")`. The cleaned "Professor Trash Wheel" data set contains `r nrow(prof_trash_wheel_df)` observations of `r ncol(prof_trash_wheel_df)` variables from `r paste(prof_trash_wheel_df$month[1], min(prof_trash_wheel_df$year), sep = ", ")` to `r paste(prof_trash_wheel_df$month[nrow(prof_trash_wheel_df)], max(prof_trash_wheel_df$year), sep = ", ")`. the cleaned "Gwynnda Trash Wheel" data set contains `r nrow(gwy_trash_wheel_df)` observations of `r ncol(gwy_trash_wheel_df)` variables from `r paste(gwy_trash_wheel_df$month[1], min(gwy_trash_wheel_df$year), sep = ", ")` to `r paste(gwy_trash_wheel_df$month[nrow(gwy_trash_wheel_df)], max(gwy_trash_wheel_df$year), sep = ", ")`.  
For the resulting datasheet, it contains `r nrow(trash_wheel_df)` observations of `r ncol(trash_wheel_df)` variables, among which the `name` variable marks the trash wheel's name `r unique(trash_wheel_df$name)  `.

It can be concluded that the total weight of trash collected by Professor Trash Wheel is `r sum(prof_trash_wheel_df$weight_tons)` and the total number of cigarette butts collected by Gwynnda in July of 2021 is 16300.



# Problem 3

```{r}
mci_baseline_df = 
  read_csv("./data_mci/MCI_baseline.csv", skip = 1, # skip the first row
           na = c(".", "NA")) |> # treat the missing value as NA
  janitor::clean_names() |> 
  mutate(
    sex = if_else(sex == 0, 'female', 'male'),
    apoe4 = case_match(
      apoe4,
      1 ~ "APOE4 carrier",
      0 ~ "APOE4 non-carrier"
    )
  ) |> 
  # the MCI onset happens after baseline
  filter(current_age < age_at_onset | is.na(age_at_onset) == 1)
```

For the data cleaning, first I skipped the first row which contains notes for the columns, and treated the missing values as NA for convenience afterwards. The `case_match` step converts the numerical values of the `sex` and `apoe4` variables to their original meanings for readability. In the end, I compared the `current_age` variable and `age_at_onset` variable to ensure the MCI onset happens after baseline. After cleaning, the data set contains  `r nrow(mci_baseline_df)` observations of `r ncol(mci_baseline_df)` variables.  
From the data set, it could be concluded that `r nrow(mci_baseline_df)` participants in total were recruited, among which `r nrow(mci_baseline_df)` participants met the criteria, and `r sum(is.na(mci_baseline_df$age_at_onset)!=1)` develop MCI. After the participants who do not meet the criteria are removed, the average baseline age is `r round(mean(mci_baseline_df$current_age), 2)`, and `r sum(mci_baseline_df$sex == "Female" & mci_baseline_df$apoe4 == "APOE4 carrier") / sum(mci_baseline_df$sex == "Female") * 100`% of women in the study are APOE4 carriers.  

### Similarly, import, clean, and tidy the dataset of longitudinally observed biomarker values.

```{r}
mci_amyloid_df = 
  read_csv("./data_mci/mci_amyloid.csv", skip = 1, # skip the first row
           na = c("NA", "Na")) |>
  janitor::clean_names() |> 
  pivot_longer(
    baseline:time_8,
    names_to = "time_in_years",
    names_prefix = "time_",
    values_to = "ABR"
  ) |> 
  mutate( # change baseline to 0 year
    time_in_years = replace(time_in_years, time_in_years == "baseline", 0),
    ABR = as.numeric(ABR)
  )
```

For the data cleaning, first, I skipped the first row, which contains notes for the columns, and treated "Na" and "NA" both as NA values because there is a "Na" value inside the "Baseline" variable. Then, I converted the five variables measuring the ratio at a given time into two variables named "time_in_years" and "ABR". After cleaning, the data set contains `r nrow(mci_amyloid_df)` observations of `r ncol(mci_amyloid_df)` variables. The ratio distribution for different years is illustrated below:

```{r}
mci_amyloid_df |> 
  ggplot(aes(y = ABR, x = time_in_years)) + 
  geom_boxplot()+
  labs(
    title = "Amyloid Beta Ratio/Years Boxplots",
   x = "Years",
    y = "Amyloid Beta Ratio")
```

From the boxplot, it could be concluded that as the number of years increases, the ratio tends to drop in overall. 

### Compare the two datasets.

Check whether some participants appear in only the baseline or amyloid datasets.

```{r}
all_participants_df =
  mci_baseline_df |> 
  full_join(mci_amyloid_df, by = join_by(id == study_id),
            keep = T) # keep both id and study_id for comparision
```

Then, `r sum(is.na(all_participants_df$id))` participants are only in the amyloid dataset, `r sum(is.na(all_participants_df$study_id))` participants are only in the baseline dataset.  
Retain only the participants who appear in both datasets.

```{r}
both_participants_df = 
  mci_baseline_df |> 
  inner_join(mci_amyloid_df, by = join_by(id == study_id))
write_csv(both_participants_df, "./data_mci/mci_result.csv") # export
```

The result contains `r nrow(both_participants_df)` observations of `r ncol(both_participants_df)` variables related to the participants' baseline demographics and amyloid $\beta$ 42/40 ratio measured every two years from the baseline to the eighth year. Among the `r length(unique(both_participants_df$id))` participants, `r sum(both_participants_df$sex == "Female") / 5` are female, `r sum(both_participants_df$apoe4 == "APOE4 carrier") / 5` are APOE4 carriers. The average current age at baseline is `r round(mean(both_participants_df$current_age), 2)` (years old), and the average length of education at baseline is `r round(mean(both_participants_df$education), 2)` years.
