---
title: "Homework2"
author: "Zhuodiao Kuang"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, echo = FALSE, cache = FALSE, message = FALSE}
rm(list = ls())
library(knitr)
## Global options
opts_chunk$set(echo      = TRUE,
	             cache     = TRUE,
               prompt    = FALSE,
               tidy      = FALSE,
               comment   = NA,
               message   = FALSE,
               warning   = FALSE,
               dpi       = 150,
               fig.width = 4, 
               fig.asp   = 5/7,
               fig.align = "center")
# attr.source = '.numberLines'
```

# Load libraries

```{r}
library(tidyverse)
library(dplyr)
library(readxl)
```


# Problem 1

We clean the 538 `pols` data, which provides information on the number of national politicians who are democratic or republican at any given time. There are some values for which `prez_gop` is `2` -- these are months in which Ford became President following Nixon's resignation. In the new `president` variable created as part of our data cleaning, we code these as `rep`` (same as values when `prez_gop` is `1`).

prez_gop: indicator of whether the president was republican on the associated date (1 = yes, 0 = no)
gov_gop: the number of republican governors on the associated date
sen_gop: the number of republican senators on the associated date
rep_gop: the number of republican representatives on the associated date
prez_dem: indicator of whether the president was democratic on the associated date (1 = yes, 0 = no)
gov_dem: the number of democratic governors on the associated date
sen_dem: the number of democratic senators on the associated date
rep_dem: the number of democratic representatives on the associated date

# First step

```{r clean_538_pols}
month_df = 
  tibble(
    month_num = 1:12,
    month_abb = month.abb,
    month = month.name
  )

pm_df = 
  read_csv("./data/pols-month.csv") |>
  separate(mon, into = c("year", "month_num", "day"), convert = TRUE) |>
  mutate(
    president = recode(prez_gop, "0" = "dem", "1" = "rep", "2" = "rep")) |>
  left_join(x = _, y = month_df) |> 
  select(year, month, everything(), -day, -starts_with("prez"))  
```


# Second step

```{r load `snp.csv`}
snp_df = 
  read_csv("./data/snp.csv") |>
  separate(date, into = c("month", "day", "year"), convert = TRUE) |>
  arrange(year, month) |>
  mutate(month = month.name[month],year = ifelse(year>=50, 1900+year, 2000+year)) |> 
  select(year, month, close) 

```

# Third step

Finally, we tidy the `unemployment` data so that it can be merged with the `pols` and `snp` datasets.

```{r clean_unemp}
unemployment = 
  read_csv("./data/unemployment.csv") |>
  rename(year = Year) |>
  pivot_longer(
    Jan:Dec, 
    names_to = "month_abb",
    values_to = "unemployment"
  ) |> 
  left_join(x = _, y = month_df) |> 
  select(year, month, unemployment)
```


# Last step


Now we merge the three datasets!

```{r merge}
data= 
  left_join(pm_df, snp_df) |>
  left_join(x = _, y = unemployment)

str(data)

```

So the pols_month dataset contains the information of presidents, governors, senators, and representatives from Republican and Democratic during different periods. 

The snp dataset contains the Standard & Poor’s stock market index during different periods. 

The unemployment dataset contains the unemployment rate in the US during different periods. 

The period of the dataset is from 1947 to 2015. The size of the merged dataset is 822 x 12.


# Problem 2

### Read and clean the Mr. Trash Wheel sheet






